A Fast kNN Algorithm Using Multiple Space-Filling Curves

### 2022-03-22 2022/Sensors/sensors.tex
fix some typos

% Keywords
\keyword{space-filling curves}

@@

\begin{itemize}
\item Brute force method. It is based on calculating all distances from the test object to other objects of the class and determining the smallest value. This method has complexity $O(n)$, where $n$ is the number of objects.
\item KD-Tree method. It is based on a special kind of binary tree, where each node represents a point in the multidimensional space. The search procedure has complexity $O(log n)$.
\item Ball-Tree method. It is yet another kind of tree structure, which has logarithmic complexity. This method is applicable to problems that have a large dimensionality.
\end{itemize}

@@

For KD-Tree, the main drawback is that it slows down when the number of objects grows, which is caused by increasing complexity of the internal tree structure. However, KD-trees have proven to be a good solution for problems with small dimensionality, which will be considered in this study.

@@

Note that in practical classification tasks, different features can have different scales, which can significantly distort the real distance between objects. Therefore it will be assumed that prior to applying kNN data scaling has been performed 
\begin{equation}
y_j \equiv \frac{y_j - y_{min}} {y_{max} - y_{min}} - \frac{1} {2}.
\end{equation}

@@

1. Divide hypercube $D$ with edge length of $1$ by coordinate hyperplanes into $2^N$ hypercubes of the first partition (with edge length of $1/2$).

@@

2. Now divide the segment $[0, 1]$ into $2^N$ equal parts, divide  each of them also into $2^N$ equal parts and so on. Denote the subinterval of the $M$-th partition by $d(M,v)$ , where $v$ is the coordinate of the left boundary point of the subinterval. Obviously, the length of the subinterval $d(M,v)$ will be equal to $2^{-NM}$. We will assume that the left boundary point belongs to the subinterval and the right boundary point does not. The only exception is the subinterval whose right boundary point is $1$, in which case it also belongs to the subinterval.

@@

Condition 1. $D(M+1,v^\prime) \in D(M,v^{\prime\prime}) \;\: i.f.f. \;\:  d(M+1,v^\prime) \in d(M,v^{\prime\prime})$.

@@

Note that the centre  $y^c(x)$ of the $M$-th partition subcube containing point $y(x)$ can be regarded as an approximation to $y(x)$ with the accuracy of $\varepsilon = 2^{-(M+1)}$, i.e.
\begin{equation}
\max_{1 \leq j \leq N} y_j^c(x)-y_j(x) \leq \varepsilon = 2^{-(M+1)}.
\end{equation}

@@

Consider a family of Peano curves
\begin{equation}
Y_L(x)=\{ y^0(x), y^1(x), ..., y^L(x) \}
\end{equation} 
instead of a single curve $y(x)$. We will construct the family of curves as follows. Let us introduce a family of hypercubes
\begin{equation}\label{hypercubes}
D_i = \{ y \in R^N: -2^{-1} \leq y_i + 2^{-l} \leq 3 \cdot 2^{-1},  1 \leq i \leq N \}, \; 0 \leq l \leq L,
\end{equation} 
where the hypercube $D_{l+1}$ is obtained by "shifting" the hypercube $D_l$ along the main diagonal by step $2^{-l}$ for each coordinate.

@@

Any subcube of the $M$-th partition of the hypercube $D_0$ will have an edge of length $2^{-(M-1)}$ and will be denoted by $D_0(M, v)$, where $v$ is the left boundary point of the subinterval $d(M, v)$ corresponding to this subcube. Then the evolvents $y^l(x)=\{y_1^l(x), ..., y_N^l(x)\}$, whose coordinates are determined by the conditions
\begin{equation}\label{maps}
y_i^l(x) = y_i^{l-1}(x) + 2^{-l}, \; 1 \leq i \leq N, \; 1 \leq l \leq L,
\end{equation} 
map the interval $[0,1]$ onto the corresponding hypercubes  $D_l, \; 1 \leq l \leq L$ (the broken line in the figure shows the image of the interval $[0,1]$, which is obtained by using the evolvent $y^0(x), x \in [0,1]$).

For any subcube $D_0(M,v)$ of the $M$-th partition of the hypercube $D_0$, there will exist a subcube $D_l(M,v)$ of the $M$-th partition of the hypercube $D_l$, and $D_l(M,v)$ can be obtained by shifting $D_0(M,v)$ along the main diagonal by the distance of $2^{-1}+2^{-2}+...+2^{-l}$.

It follows from formulae (\ref{hypercubeD0}) and (\ref{maps}) that if an interval $d(M,v)$ is mapped onto the subcube $D_0(M,v)$, then there exists a family of subcubes
\begin{equation}
D_l(M,v) = y^l(d(M,v)), 1 \leq l \leq L,
\end{equation} 
connected with the corresponding subintervals $d_l(M,v_l) \subset [0,1]$, where $d(M,v) = d_0(M,v_0), v_0 = v$, such that 
\begin{equation}
D_l(M,v) = y^l(d_l(M,v)), 1 \leq l \leq L,
\end{equation}

@@

As an illustration, Figure \ref{fig2} shows a family of space-filling curves in extended domains $D_l$, where a square with a dark border highlights the hypercube $D$ belonging to each extended domain. The dots indicate proximate objects in the domain $D$, the distance between which will vary depending on the one-dimensional scale used. The effect of maintaining distance proximity on one of the scales is clearly visible. For example, objects marked with green dots will be far away on the first two one-dimensional scales and close to each other on the third one. Objects marked in blue will be far away on the first and third scales and close to each other on the second scale.

@@

The degree of proximity of a test object to one of the known objects can be expressed in terms of the distance on the $l$-th scale.
\begin{equation}
d_i^n = |x_i^n - x^{*n}|
\end{equation}

Thus, the distance $d(\omega_i,\omega_j)$ between two objects $\omega_i$ and $\omega_j$ on a one-dimensional scale will correspond to the minimum distance among all the mappings. If we find the minimum distance among all these distances, we get the nearest neighbour
\begin{equation}
d_{min} = \min \{ d_i^n \} = \min \{ |x_i^n - x^{*n}| \}.
\end{equation}

Next, the search procedure needs to be run as many times as there are nearest neighbours to be identified. The main problem is that an object previously found on one scale can be found again, including on another scale. This requires either removing the values of x found objects from all scales or searching for $K$ nearest neighbours within each scale and then merging the results.

>>>>

% Keywords
\keyword{machine learning, kNN, dimensionality reduction, multiple space-filling curves}

@@

\begin{itemize}
\item Brute force method. It is based on calculating all distances from the test object to other objects of the class and determining the smallest value. This method has complexity $O(n)$, where $n$ is the number of objects.
\item KD-tree method. It is based on a special kind of binary tree, where each node represents a point in the multidimensional space. The search procedure has complexity $O(log(n))$.
\item Ball-tree method. It is yet another kind of tree structure, which has logarithmic complexity. This method is applicable to problems that have a large dimensionality.
\end{itemize}

@@

For kd-tree, the main drawback is that it slows down when the number of objects grows, which is caused by increasing complexity of the internal tree structure. However, kd-trees have proven to be a good solution for problems with small dimensionality, which will be considered in this study.

@@

Note that in practical classification tasks, different features can have different scales, which can significantly distort the real distance between objects. Therefore it will be assumed that prior to applying kNN data scaling has been performed 
\begin{equation}
y_j = \frac{y_j - y_{min}} {y_{max} - y_{min}} - \frac{1} {2}.
\end{equation}

@@

1. Divide hypercube $D$ form (\ref{D}) with edge length of $1$ by coordinate hyperplanes into $2^N$ hypercubes of the first partition (with edge length of $1/2$).

@@

2. Now divide the segment $[0, 1]$ into $2^N$ equal parts, divide  each of them also into $2^N$ equal parts and so on. Denote the subinterval of the $M$-th partition by $d(M,v)$, where $v$ is the coordinate of the left boundary point of the subinterval. Obviously, the length of the subinterval $d(M,v)$ will be equal to $2^{-NM}$. We will assume that the left boundary point belongs to the subinterval and the right boundary point does not. The only exception is the subinterval whose right boundary point is $1$, in which case it also belongs to the subinterval.

@@

Condition 1. $D(M+1,v^\prime) \in D(M,v^{\prime\prime})$ i.f.f. $d(M+1,v^\prime) \in d(M,v^{\prime\prime})$.

@@

Note that the centre  $y^c(x)$ of the $M$-th partition subcube containing point $y(x)$ can be regarded as an approximation to $y(x)$ with the accuracy of $\varepsilon = 2^{-(M+1)}$, i.e.
\begin{equation}
\max_{1 \leq j \leq N} \left| y_j^c(x)-y_j(x) \right| \leq \varepsilon = 2^{-(M+1)}.
\end{equation}

@@

Consider a family of Peano curves
\begin{equation}
Y_L(x)=\{ y^0(x), y^1(x), ..., y^L(x) \}
\end{equation} 
instead of a single curve $y(x)$. We will construct the family of curves as follows. Let us introduce a family of hypercubes
\begin{equation}\label{hypercubes}
D_i = \{ y \in R^N: -2^{-1} \leq y_i + 2^{-l} \leq 3 \cdot 2^{-1},  1 \leq i \leq N \}, \; 0 \leq l \leq L,
\end{equation} 
where the hypercube $D_{l+1}$ is obtained by shifting the hypercube $D_l$ along the main diagonal by step $2^{-l}$ for each coordinate.

@@

Any subcube of the $M$-th partition of the hypercube $D_0$ will have an edge of length $2^{-(M-1)}$ and will be denoted by $D_0(M, v)$, where $v$ is the left boundary point of the subinterval $d(M, v)$ corresponding to this subcube. Then the evolvents $y^l(x)=\{y_1^l(x), ..., y_N^l(x)\}$, whose coordinates are determined by the conditions
\begin{equation}\label{maps}
y_i^l(x) = y_i^{l-1}(x) + 2^{-l}, \; 1 \leq i \leq N, \; 1 \leq l \leq L,
\end{equation} 
map the interval $[0,1]$ onto the corresponding hypercubes  $D_l, \; 1 \leq l \leq L$ (the broken line in Fig.~\ref{fig2} shows the image of the interval $[0,1]$, which is obtained by using the evolvent $y^0(x), x \in [0,1]$).

For any subcube $D_0(M,v)$ of the $M$-th partition of the hypercube $D_0$, there will exist a subcube $D_l(M,v)$ of the $M$-th partition of the hypercube $D_l$, and $D_l(M,v)$ can be obtained by shifting $D_0(M,v)$ along the main diagonal by the distance of 
\[
2^{-1}+2^{-2}+...+2^{-l}.
\]

It follows from formulae (\ref{hypercubeD0}) and (\ref{maps}) that if an interval $d(M,v)$ is mapped onto the subcube $D_0(M,v)$, then there exists a family of subcubes
\begin{equation}
D_l(M,v) = y^l(d(M,v)), 1 \leq l \leq L,
\end{equation} 
connected with the corresponding subintervals $d_l(M,v_l) \subset [0,1]$, where $d(M,v) = d_0(M,v_0), v_0 = v$, such that 
\begin{equation}
D_l(M,v) = y^l(d_l(M,v)), 1 \leq l \leq L.
\end{equation}

@@

As an illustration, Fig. \ref{fig2} shows a family of space-filling curves in extended domains $D_l$, where a square with a dark border highlights the hypercube $D$ belonging to each extended domain. The dots indicate proximate objects in the domain $D$, the distance between which will vary depending on the one-dimensional scale used. The effect of maintaining distance proximity on one of the scales is clearly visible. For example, objects marked with green dots will be far away on the first two one-dimensional scales and close to each other on the third one. Objects marked in blue will be far away on the first and third scales and close to each other on the second scale.

@@

The degree of proximity of a test object to one of the known objects can be expressed in terms of the distance
\begin{equation}
d_i^l = |x_i^l - x^{*l}|
\end{equation} 
on the $l$-th scale.

Thus, the distance $d(\omega_i,\omega_j)$ between two objects $\omega_i$ and $\omega_j$ on a one-dimensional scale will correspond to the minimum distance among all the mappings. If we find the minimum distance among all these distances, we get the nearest neighbour
\begin{equation}
d_{min} = \min \{ d_i^l \} = \min \{ |x_i^l - x^{*l}| \}.
\end{equation}

Next, the search procedure needs to be run as many times as there are nearest neighbours to be identified. The main problem is that an object previously found on one scale can be found again, including on another scale. This requires either removing the values of $x$ found objects from all scales or searching for $K$ nearest neighbours within each scale and then merging the results.
